---
title: "CSSS 563 - HW1"
author: "Courtney Allen"
date: "4/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(SciViews)
library(kableExtra)
```


## Question 1

Suppose the instantaneous mortality rate for a population is a constant that does not
depend on age, equal to 0.02 per year.


### Q1.A

**Find analytically and plot the probability density function of age at death for this population.**


Probability density function is:

$$
 f(x) = \lambda e ^{-\lambda x}
$$
$$
\lambda = 0.02
$$
$$
f(x) = 0.02 e ^{-0.02 x}
$$

The probability distribution of age at death is exponential: 

```{r}
lambda <- 0.02
beg_age <- 0
end_age <- 110
population_ages <- (tibble(age = seq(beg_age,end_age, 1)))


pdf = function(x) lambda*exp(-lambda*x)
pdf_plot = 
  population_ages %>% 
  mutate(PDF = pdf(age)) %>% 
  ggplot(aes(x=age, y=PDF)) + 
  geom_line() + 
  theme_classic() + 
  labs(
    title = "Probability Density Function", 
    x = "Age (years)",
    y = "Probability"
    )

pdf_plot

```




### Q1.B

**Find the probability that a member of this population is still alive at age 70.**

The probability that a member of the population is still alive is represented by the Survival Function:

$$
S(x) = Pr[X>x] = 1-F(x) = e ^{-\lambda x}
$$
$$
S(x) = e ^{-0.02*70}
$$
```{r}

survival = function(x) exp(-lambda*x)

sx_70 <- survival(70)
```

The probablity of a member of the population surviving to age 70 is **`r round(sx_70,2)`**.

### Q1.C

**Find the probability that a member of this population dies before age 6.**

The probability that a member of the population dies by a certain age is represented by the Cumulative Distribution Function:

$$
 f(x) = 1- e ^{-\lambda x}
$$
$$
 f(x) = 1- e ^{-0.02 * 6}
$$

```{r}

cdf = function(x) 1 - exp(-lambda*x)

qx_6 <- cdf(6)

```

The probability of a member of the population surviving to age 70 is **`r round(qx_6,2)`**.


### Q1.D

**Find the life expectancy at birth for a member of this population.**

Life expectancy can be found by integrating the survival function from 0 to infinity.

$$
e_0 = E(X) = \int_{0}^{\infty} S(u)du
$$

```{r}
  


e0 <- 1/0.02
  

```

The life expectancy at birth is **`r round(e0,2)`**.

### Q1.E

**Find the life expectancy at age 50 for a member of this population.**

Life expectancy at age 50 is: 

$$
e_0 = E(X) = \int_{50}^{\infty} \frac{S(u)du}{S(50)}
$$

```{r}
survival_function <- function(x) {
  exp(-lambda*x)
}
e50_temp <- integrate(survival_function, lower = 50, upper = Inf)
e50 <- e50_temp$value/survival_function(50)

```


The life expectancy at age 50 is **`r round(e50,2)`**.


### Q1.F

**Find the median age at death for this population**

```{r}

population_ages <- (tibble(age = seq(1,110, 1)))

med_death <- (ln(2))/lambda
```


The median age at death is `r round(med_death,2)`.


## Question 2

**Suppose the instantaneous mortality rate at age x for a cohort is$\mu$(x) = (0.0168x2 -  0.668x + 8)/1000**

**Denote by X the age at death of a person randomly chosen from this cohort (a random variable).**

### Q2.A

**Plot the instantaneous mortality rate for this cohort against age.**

```{r}
hazard_fun = function(x) (0.0168*x^2 - 0.668*x + 8)/1000
imr_plot <- population_ages %>%
  mutate(mortality = hazard_fun(age)) %>%
  ggplot(aes(x = age, y = mortality)) +
  geom_line() +
  theme_bw() +
  theme(text = element_text(family = "serif")) +
  labs(
    title = "Mortality Rate vs Age",
    x = "Age (years)",
    y = "Instantaneous Mortality Rate"
  )
imr_plot

```



### Q2.B

**Find analytically and plot the cumulative hazard function of this distribution, (x).**

The cumulative hazard function is $\Lambda(x) = \int_{0}^{x} \mu(u)du$.


$$
\begin{aligned} \Lambda(x) &= \int_{0}^{x} \mu(u) du \\
&= \int_{0}^{x} \left[(0.0168x^2 − 0.668x + 8)/1000 \right]dx \\ 
&= \int_{0}^{x} \dfrac{21x^2-835x+10000}{1250000}dx\\
\end{aligned}
$$


```{r}

cum_hazard_fun <- function(x) (-835*x + 21*x^2 + 10000)/1250000

chf_plot <- ggplot(population_ages, aes(x = age, y = cum_hazard_fun(age))) +
  geom_line() +
  theme_classic() +
  labs(
    title = "Cumulative Hazard vs Age",
    x = "Age (years)",
    y = "Cumulative Hazard"
  )

chf_plot



```

### Q2.C

**Find analytically and plot the survival function, S(x).**

The survival function is: 

$$
S(x) = exp^{-1*cumulativehazard}
$$

```{r}

survival_fun <- function(x) exp(-1 * cum_hazard_fun(x))
survival_fun <- function(x) exp(-1 * ((0.0168*x^2 - 0.668*x + 8)/1000))
surv_plot = 
  population_ages %>%
  mutate(survive = survival_fun(age)) %>%
  ggplot(aes(x=age, y=survive)) + 
  geom_line() + 
  theme_classic() + 
  labs(
    title = "Survival Function", 
    x = "Age (years)",
    y = "Probability of surviving to age X"
    )

surv_plot

```

### Q2.D

**Find and plot the probability density function of X, f(x).**

```{r}

pdf_fun <- function(x) survival_fun(x) * hazard_fun(x)
pdfun_plot <-
  population_ages %>%
  mutate(pdf = pdf_fun(age)) %>%
  ggplot(aes(x = age, y = pdf)) +
  geom_line() +
  theme_classic() + 
  labs(
    title = "Probability density vs Age",
    x = "Age (years)",
    y = "Probability density")

pdf_plot


```

### Q2.E

**Find the life expectancy at birth of a member of this cohort.**

Life expectancy is: 

$$
e_0 = E(X) = \int_{0}^{\infty} S(u)du
$$


```{r}

q2_e0 <- integrate(survival_fun, lower = 0, upper = Inf)

```

The life expectancy at birth is **`r round(q2_e0$value,2)`**.


### Q2.F

**Find the life expectancy at age 10 of a member of this cohort.**

Life expectancy at age 10 is: 

$$
e_{10} = E(X) = \int_{10}^{\infty} \frac{S(u)du}{S(10)}
$$

```{r}
e10_temp <- integrate(survival_function, lower = 10, upper = Inf)
e10 <- e10_temp$value/survival_function(10)

```

The life expectancy at birth is **`r round(e10,2)`**.

### Q2.G

**Find 45q15 for this cohort.**

$$
_{n}q_{x} = \frac{_{n}d_{x}}{l_{x}} = \frac{l_{x}-l_{x+n}}{l_{x}} = \frac{S_{(x)}-S_{(x+n)}}{S_{(x)}}
$$

```{r}

nqx <- function(x, n) (survival_fun(x) - survival_fun(x + n)) / survival_fun(x) 
q45_15 <- round(nqx(15, 45), 3)

```

The probability of dying between ages 15 and 60 is **`r round(q45_15,2)`**

## Question 3

**From the UN's 2019 World Population Prospects extract and show:**

```{r}
library(wpp2019)
library(wppExplorer)

```

### Q3.A

**estimates of the male population of Peru in each of the age groups 70-74, 75-79, 80-84, 85-89, 90-94, 95+ in 2015 and 2020**

```{r}

# download estimates of males
data("popM")

# filter for Peru, ages 70+
df_pe <- popM %>%
  filter(country_code == 604, 
         age == c("70-74", "75-79", "80-84", "85-89", "90-94", "95-99", "100+")) %>%
    select("age", "2015", "2020")

# make a table
df_pe %>%
  filter(age!="100+") %>% 
  kable() %>%
  kable_paper("hover", full_width = F)

```

### Q3.B

**estimates of the number of deaths in each of these age groups during the period 2015-2020.**

Deaths can be estimated by:

$$
_{n}d_{x} = l_{x} - l_{x+n}
$$

```{r}

# create deaths by measuring lx - lx+n
df_pe <- df_pe %>% 
  mutate(deaths = `2015` - lead(`2020`))

# make a table
df_pe %>%
  filter(age!="100+") %>% 
  rename(`deaths (thousands)` = deaths) %>%
  kable() %>%
  kable_paper("hover", full_width = F)

```

## QUESTION 4



### Q4.A

**Use these estimates to:**
**estimate $_5M_x$ for ages x = 70; 75; 80; 85; 90; 95 for males in Peru for the period 2015-2020**

The age-specific mortality rate is:

$$
_{n}M_{x} = \frac{_{n}d_{x}}{_{n}L_{x}}
$$



```{r}

df_pe <- df_pe %>%
  mutate(nLx = (5*(`2015`+`2020`)/2),
         nMx = deaths/nLx)


df_pe %>%
  filter(age!="100+") %>% 
  kable() %>%
  kable_paper("hover", full_width = F)

```

### Q4.B

**estimate 5qx for ages x = 70; 75; 80; 85; 90; 95 for males in Peru for the period 2015-2020, using each of the two approximations discussed in class.**



Solution 1:

$$
_nq_x \approx 1-exp(-n_nM_x)
$$
```{r}

# estimate first method of approximation for nqx
df_pe <- df_pe %>%
  mutate(nqx_1 = 1-exp(-5*nMx))


```



Solution 2:

$$
_nq_x \approx \frac{n_nM_x}{1+n {_nM_x}/2}
$$


```{r}

# estimate second method of approximation for nqx
df_pe <- df_pe %>%
  mutate(nqx_2 = (5*nMx)/(1+5*nMx/2))

df_pe %>%
  filter(age!="100+") %>%
  select(age, `2015`, `2020`, deaths, nMx, nqx_1, nqx_2) %>%
  kable() %>%
  kable_paper("hover", full_width = F)
```





### Q4.C

**estimate $_5m_x$ for the same ages**

The age-specific mortality rate is:

$$
_{n}m_{x} = \frac{_{n}d_{x}}{_{n}L_{x}} = \frac{_{n}d_{x}}{nl_{x} - \frac{1}{2}n_nd_x}
$$



```{r}

df_pe <- df_pe %>%
  mutate(nLx_approx = (5*`2015`-(1/2)*5*deaths),
         nmx = deaths/nLx_approx)


df_pe %>%
  filter(age!="100+") %>% 
  select(age, `2015`, `2020`, deaths, nLx_approx, nmx) %>%
  kable() %>%
  kable_paper("hover", full_width = F)

```

### Q4.D

**Compare the estimates from the two approximations. Which of the two approximations do you think is better? Why?**

```{r}

df_pe %>%
  filter(age!="100+") %>% 
  select(age, `2015`, `2020`, deaths, nMx, nqx_1, nqx_2) %>%
  kable() %>%
  kable_paper("hover", full_width = F)

```

The table above shows the two approximations for $_nq_x$, the probability of dying in an age group of those aged x to x+n. The estimates are very close, though as $_nM_x$ gets larger, there is a larger discrepancy between the approximations. 

I believe the first approximation ($_nq_x \approx 1-exp(-n_nM_x)$) is better because the equation is based on the assumption that the mortality rate is constant over time within the age interval. I believe this is intuitive and *mostly* true for these age groups, except not for the oldest age groups.

*from graded notes* second approx is closer to wpp reports.     

### Q4.E

**Compare the estimates of $_5M_x$ and $_5m_x$  with one other, and comment on any dfferences.**

```{r}
df_pe %>%
  filter(age!="100+") %>% 
  select(age, `2015`, `2020`, deaths, nLx, nLx_approx, nMx, nmx) %>%
  kable() %>%
  kable_paper("hover", full_width = F)
```


For each age group $_nM_x$ is smaller than $_nm_x$. $_nM_x$ uses the person years calculated in $_nL_x$ column, while the $_nm_x$ utilizes the `nLx_approx` column (which assumes linear function of person years over the interval). $_nM_x$ may be underestimating mortality by using person-years estimates from population counts, while $_nm_x$ actually accounts for deaths in the cohort when estimating person-years. As can be seen, $_nL_x$ estimates more person-years lived than `nLx_approx`. Older folks who die in these age groups may not live half of the interval as is assumed, so $_nm_x$ may be more of an accurate estimate of the mortality rate.

## QUESTION 5

**For each of the two approximations:**

### Q5.A

**Derive a life table ($l_x$ only) for a cohort of 10,000 males aged 70.**

To construct a cohort life table *ONLY* using nqx values we'll rely on the following equations:


**note from graded version** 
5dx = 5qx*lx

life expectance e70 = $\sum^95_{i=70} * 5Li$ over l70
Deaths
$$
_nd_x = _nq_x*_nlx
$$

Probability of surviving
$$
_np_x = 1-_nq_x
$$

Person-years
$$
_nLx = \frac{_{n}d_{x}}{nl_{x} - \frac{1}{2}n_nd_x}
$$

Person-years lived above age x
$$
_nT_x = \sum_{a=x}^{inf}{_nL_a}
$$

Life expectancy
$$
e^0_x = \frac{T_x}{l_x}
$$




```{r life_table_approx_1}

lt_1 <- df_pe %>%
  select(age, nqx_1) %>%
  mutate(
    n = 5,
    npx = 1-nqx_1,
    nqx = case_when(
      age=="100+" ~ 1,
      TRUE ~ nqx_1    )) %>%
  select(-nqx_1)


# rest of ages
x <- seq(70,100,5)
radix <- 10000
lx <- radix
npx <- lt_1$npx

#fill out rest of table
for (a in x[-length(x)]) {
  l <- lx[which(x == a)] * npx[which(x == a)]
  lx <- c(lx, l)
}
lt_1 <- lt_1 %>%
  mutate(
    x = x,
    lx = lx,
    ndx = lx*nqx,
    Lx = n*lx - n*ndx/2,
    Tx = rev(Lx) %>% coalesce(0) %>% cumsum() %>% rev(),
    ex = Tx / lx
  )

```



```{r life_table_approx_2}

lt_2 <- df_pe %>%
  select(age, nqx_2) %>%
  mutate(
    n = 5,
    npx = 1-nqx_2,
    nqx = case_when(
      age=="100+" ~ 1,
      TRUE ~ nqx_2    )) %>%
  select(-nqx_2)


# rest of ages
x <- seq(70,100,5)
radix <- 10000
lx <- radix
npx <- lt_2$npx

#fill out rest of table
for (a in x[-length(x)]) {
  l <- lx[which(x == a)] * npx[which(x == a)]
  lx <- c(lx, l)
}
lt_2 <- lt_2 %>%
  mutate(
    x = x,
    lx = lx,
    ndx = lx*nqx,
    Lx = n*lx - n*ndx/2,
    Tx = rev(Lx) %>% coalesce(0) %>% cumsum() %>% rev(),
    ex = Tx / lx
  )


```


Below are two life tables, each life table uses a different approximation technique for nqx.

```{r}

lt_1 %>% select(-x,-n) %>%
  kable %>%
  kable_paper("hover", full_width = F)

lt_2 %>% select(-x,-n) %>%
  kable %>%
  kable_paper("hover", full_width = F)

```

### Q5.B

**Estimate the life expectancy at age 70 from each of the two approximations, and compare the two estimates. Comment on any differences, and why they arise.**

```{r}


cbind(age = lt_1$x, nqx_1=lt_1$nqx, nqx_2=lt_2$nqx, ex_1= lt_1$ex, ex_2=lt_2$ex) %>%
  kable(., digits=2)  %>%
  kable_paper("hover", full_width = F)


```

Above, you can compare the life expectancies that use different $_nq_x$ approximations. 

The first approximation technique estimates slightly higher life expectancies. These are very slight and most life expectancies are within a few decimal points. The first approximation technique estimates slightly lower probabilities of dying which is why the life expectancies are slightly longer for this approach.






# Appendix

```{r getlabels, include=FALSE}
labs <- knitr::all_labels()
labs <- labs[!labs %in% c("setup", "toc", "getlabels", "allcode")]
```

```{r allcode, ref.label=labs, eval=FALSE, echo=TRUE}
```


